---
title: "CM515 Day 1: Plotting with ggplot()"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(datasets)
library(tidyverse)
library(knitr)
library(ggplot2)
library(RColorBrewer)
library(scales)
```

# 1. Basic Plot Functions

------------------------------------------------------------------------

R has a built-in basic plot function. However it is pretty limited in
scope.

```{r, echo=T, message=F, warning=F}
####
#01#
####
plot(c(1,2,3,4,5),c(1,2,3,4,5), xlab="X-Axis", ylab="Y-Axis", type='l', main="This is a Graph", col='blue')




```

# 2. data()

------------------------------------------------------------------------

Before I welcome you to the future with ggplot, we need some data.

A ton of datasets can be accessed with data().

```{r, echo=T, message=F, warning=F}
####
#02#
####
data()

```

# 2b. data()

------------------------------------------------------------------------

We will use the iris data set primarily.

It is worth getting familiar with the data.

```{r, echo=T, message=F, warning=F}
####
#02b#
####

?iris

head(iris)
summary(iris)


```

# 3. ggplot()

------------------------------------------------------------------------

Now let's try to graph, just using ggplot().

...well?

Nothing happens because we did not specify aesthetics or a geometry.

```{r, echo=T, message=F, warning=F}
####
#03#
####
iris %>%
  ggplot()

```

------------------------------------------------------------------------

# 3b. Mapping x and y Aesthetics with aes() Inside of ggplot()

------------------------------------------------------------------------

We're closer now but we didn't specify a geometry.

```{r, echo=T, message=F, warning=F}
####
#03b#
####
iris %>%
  ggplot(aes(#put in a value for x and y based on the available columns))

```

------------------------------------------------------------------------

# 4. Specifying a Geometry

------------------------------------------------------------------------

Specifying aesthetics and a geometry are the minimum requirement to make
a plot using ggplot(). In this case, I specified geom_point(). Notice
that I added it to the ggplot() function with "+". "+" in this case acts
like %\>% for all code involving ggplot.

```{r, echo=T, message=F, warning=F}
####
#04#
####
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_line


```

------------------------------------------------------------------------

# 4b. More about Geometry

------------------------------------------------------------------------

Each geometry has its own unique aesthetics and content. The aesthetics
of a scatter plot require x and y coordinates. This scatter plot allows
us to modify:

size=# (for the size of points) color="" (for the color) shape= "" (for
the shape of the points) alpha= a range of 0-1 alpha is transparency
where 0 is transparent and 1 is visible (great for overlapping points)

Play around with color, shape, fill, and stroke. See what you like.
Shapes can be strings or numbers.

```{r, echo=T, message=F, warning=F}
####
#04b#
####


# show them 

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(#try customizing your points)

```

# 5. Coloring by Factor

------------------------------------------------------------------------

We can color points by a factor, which is typically a categorical
variable, to highlight specific trends. This is done by specifying
"color" (sometimes abbreviated as "col") within the aes() function. In
this case, we specify that the points should be colored according to the
factor "Species".

Note: For some geometries and/or variables you may need to use "fill"
instead of "color" but overall they function similarly.

```{r, echo=T, message=F, warning=F}
####
#08#
####


iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Sepal.Length)) +
  geom_point() 

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() 


```

------------------------------------------------------------------------

# 5b. Using Shape to Denote a Factor

------------------------------------------------------------------------

Similar to "color" we can also make points different shapes based on a
factor. To do so, we specify "shape = Species" inside the aes()
function. By specifying "color = Species" and "shape = species", each
species gets a unique shape and color.

```{r, echo=T, message=F, warning=F}
####
#05b#
####
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, shape = Species)) +
  geom_point() 

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) +
  geom_point()

```

# 5c. Manipulating Geometry with Aesthetics

------------------------------------------------------------------------

Because geometry and aesthetics are somewhat intertwined, you can factor
certain geometric components in your mapping.

Size and color do not have to be limited to specifics, but can also be
linked to expressions

Note that ggplot automatically creates legends for your specification in
mapping.

```{r, echo=T, message=F, warning=F}
####
#05c#
####

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Sepal.Width > 3, shape=Species, size=Sepal.Length > 7)) +
  geom_point()


```

# 5d. Geometries vs Aesthetics

------------------------------------------------------------------------

It is important to note that you can customize your graph in both
mapping and geometry, but there are advantages to each.

Geometry lets you customize your entire mapping. Aesthetics lets you
customize your mapping with respect to your data.

Depending on what you are trying to do, you may want your mapping in one place or the other.

Put a color in aesthetics and a color in your geometry. Which color did you get? Does one argument override the other?

```{r, echo=T, message=F, warning=F}
####
#05d#
####
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()

```

------------------------------------------------------------------------

# 6. Plotting a Continuous Variable Against A Categorical Variable

------------------------------------------------------------------------

## As a Scatter Plot

```{r, echo=T, message=F, warning=F}
####
#6#
####
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_point(size=3) 
```

Some of the points are overlapping. We can try using alpha here for
transparency.

```{r, echo=T, message=F, warning=F}

iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_point(size = 3, 
             alpha = .5) 

```

## As a Jitter Plot

It looks like we have reached the limitations of a scatter plot for this
context.

We will need to explore another geometry. geom_jitter spaces out
overlapping points.

Now we can see individual points better.

```{r, echo=T, message=F, warning=F}

iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_jitter(size = 3,
              alpha = .5)

```

The default "jitter" settings "spread" the points too much for my
liking, so I always adjust them using "width" and "height" within the
geom_jitter() function.

```{r, echo=T, message=F, warning=F}

iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_jitter(size = 3,
              alpha = .5, 
              width = .2, 
              height = 0) 

```

## As a Box Plot

We could also use geom_boxplot().

```{r, echo=T, message=F, warning=F}

iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_boxplot() 

```

## As a Bar Plot with Error Bars

We could use geom_col() and geom_errorbar() to make a bar plot but this
requires calculating the Species mean and standard error "by hand" using
the group_by() and summarise() functions and then plotting with
ggplot().

Note that we must specify "ymin" and "ymax" within the aes() of
geom_errorbar(). These are the lower and upper bounds of the error bars.
I also adjusted the width of the error bars by specifying "width = .1"
within geom_errorbar() but outside of aes().

```{r, echo=T, message=F, warning=F}

iris_2.0 <- iris %>%
  group_by(Species) %>%
  summarise(mean = mean(Sepal.Length), 
            sd = sd(Sepal.Length), 
            n = n(), 
            se = sd/sqrt(n))
iris_2.0

iris_2.0 %>%
  ggplot(aes(x = Species, y = mean, ymax = mean + se, ymin = mean - se, fill = Species)) +
  geom_col() +
  geom_errorbar(width=0.1)
```

Specifying "fill" instead of "color" has a similar but different effect
on the plot.

Notice here color relates to the outer boundary only.

```{r, echo=T, message=F, warning=F}

iris %>%
  group_by(Species) %>%
  summarise(mean = mean(Sepal.Length), 
            sd = sd(Sepal.Length), 
            n = n(), 
            se = sd/sqrt(n)) %>%
  ggplot(aes(x = Species, y = mean, color = Species)) +
  geom_col() +
  geom_errorbar(aes(ymax = mean + se, ymin = mean - se), 
                width = .1) 

```

------------------------------------------------------------------------

# 7. Overlaying Geoms: Boxplot with Points Overlain

------------------------------------------------------------------------

Can you add multiple plots over each other?


```{r, echo=T, message=F, warning=F}
####
#07#
####
  iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_jitter(width = .2, 
              height = 0,
              alpha = .5, 
              size = 3) 

```
Of course you can! However,
because geom_boxplot() plots outliers as points geom_jitter plots all
points, some observations are plotted twice. This problem can be seen
above for the vriginica species. To correct this, we can specify "outlier.shape = NA" withing
geom_boxplot() as seen below.

```{r, echo=T, message=F, warning=F}

iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 3,
              width = .2,
              height = 0,
              alpha = .5)

```

# 8. Positioning Data

What do we do if we have overlapping data?

There are two ways we can resolve this issue.

In bar plots, this is easily solved by position arguments. If we graph a
count of how many leaves were a specific sepal length, the counts
overlap between species.

```{r, echo=T, message=F, warning=F}

####
#08#
####

iris %>%
  ggplot(aes(x=Sepal.Width, fill=Species))+
  geom_bar()

```

We can re-position the data using position_dodge.

```{r, echo=T, message=F, warning=F}

install.packages("ggridges")
library(ggridges)
iris %>%
  ggplot(aes(x=Sepal.Width, fill=Species))+
  geom_bar( position=position_dodge())

#this separates the columns nicely, but they are not all the same size.
#let's try changing the width argument,

iris %>%
  ggplot(aes(x=Sepal.Width, fill=Species))+
  geom_bar( position=position_dodge(), width=1)

#woah that is too big!


iris %>%
  ggplot(aes(x=Sepal.Width, fill=Species))+
  geom_bar( position=position_dodge(preserve='single'))

#much better!


#we could use a geom-density plot alternatively

iris %>%
  ggplot(aes(x=Sepal.Width, fill=Species))+
  geom_density(alpha=0.5)


iris %>%
  ggplot(aes(x=Sepal.Width, y=Species, fill=Species))+
  geom_density_ridges(alpha=0.5)

```

# 9. Adding Labels and a Title

------------------------------------------------------------------------

Adding the labs() function with "+" lets us give the plot x and y labels
as well as a title and subtitle

You can also change the label for any component that has a built in
legend.

Come up witb some labels!

```{r, echo=T, message=F, warning=F}
####
#09#
####

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Sepal.Width > 3)) +
  geom_point() +
  labs(x = "", y = "", 
       title = "", 
       subtitle= "",
       color= "")

#note- using "" for a value in labs is an easy trick if you want to have no label for an axis, title, or legend title

```

------------------------------------------------------------------------

# 10. Centering the Title

------------------------------------------------------------------------

Many parameters can be adjusted in the theme function. Overall, its
fairly unintuitive but typically, someone has done what you would like
to do and posted the solution on the internet so it's worth looking it
up. I never remember how to center the title so I always have to look it
up and copy the code.

```{r, echo=T, message=F, warning=F}
####
#10#
####
iris %>%
  ggplot() +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, color=Sepal.Width > 3)) +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length", 
       subtitle= "For Virginica, Versicolor, and Setosa",
       color= "Large Sepal Length") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) 

```

------------------------------------------------------------------------

# 11. Choosing a Theme

------------------------------------------------------------------------

There are many themes in ggplot(). The theme can be added using "+" and
specifying a theme, such as "theme_bw()". 

There are many many many more themes available than just those listed.

ggthemes has a ton more options

It goes beyond the scope of the class, but one can even generate their
own themes. You can also meet halfway, use a theme that you like, and then build off of it.

Look through the themes and ggthemes to find what you like best.

```{r, echo=T, message=F, warning=F}
####
#11#
####

#install.packages("ggthemes")
library(ggthemes)

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 

#There are A LOT of things you can do with theme

```

------------------------------------------------------------------------

# 12. Scales and Coordinates

------------------------------------------------------------------------

Scales

Scales allow us to define our mapping.

We can use scale_color_gradient or scale_color_brewer to take a
continuous variable and make a color scheme that is both easy and
pleasant on the eyes.

However, scale_color_brewer is only for categorical variables, not
continuous ones.

You can also use scale_color_manual to make your own colorset.

(There is also a scale_shape_manual, but I will let you figure that one
out ;)

To make a colorblind friendly graph, you need only put in at the end of
your graph the function: scale_colour_colorblind()

```{r, echo=F, message=F, warning=F}
####
#12#
####


display.brewer.all()

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_stata() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2")

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_stata() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(breaks=c("setosa", "versicolor", "virginica"), values=c("red", "yellow3", "blue"))

iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_stata() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_colour_colorblind()
```

```{r, echo=F, message=F, warning=F}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_stata() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(breaks=c("setosa", "versicolor", "virginica"), values=c("goldenrod2", "steelblue4", "chocolate4"))
```

```{r, echo=F, message=F, warning=F}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color=Sepal.Width)) +
  geom_point(size=2) +
  labs(x = "Sepal Length", y = "Sepal Width", 
       title = "Sepal Width vs. Sepal Length") +
  theme_dark() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_gradient(low = "yellow", high = "red")
```

# 12b. Scales for your axis

------------------------------------------------------------------------

Scales

You can also use scales to modify your axis. If we graph petal length
and petal width, the standard x axis looks very sparse. What if we don't
care about the setosa data and are more interested in the spread of
versicolor and virginica data?

We can use scales to zoom in on that data and provided more detailed
tick marks to tease out the details.

We can use seq as a shorter way of making an axis: 1st number-lowest
value on axis 2nd number-highest value on axis 3rd number-make tick
marks every n times

You can also set no tick marks

Try to scale the graph to just show the data for setosa.

```{r, echo=F, message=F, warning=F}
####
#12b#
####

iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2")

iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2") +
  scale_x_continuous (breaks=c(0, 1, 2, 3, 4, 5, 6, 7), limits = c(0,7)) +
  scale_y_continuous(limits=c(0,2.5))

iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2") +
  scale_x_continuous (name= "Petal Length (cm)", breaks=c(3.0, 3.5, 4.0, 4.5, 5.0, 5.5, 6.0, 6.5, 7.0),limits = c(3,7)) +
  scale_y_continuous(name="Petal Width (cm)", breaks = c(1.0, 1.25, 1.50, 1.75, 2.0, 2.25, 2.50), limits=c(1.0,2.5))

iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2") +
  scale_x_continuous (name= "Petal Length (cm)", breaks=seq(3.0, 7.0, 2.0),limits = c(3,7)) +
  scale_y_continuous(name="Petal Width (cm)", breaks = seq(1.0, 2.5, 0.25), limits=c(1.0,2.5))

#you can also set breaks to null to remove tick marks all together
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_brewer("Species", palette="Dark2") +
  scale_x_continuous (name= "Petal Length (cm)", breaks=seq(3,7,2),limits = c(3,7)) +
  scale_y_continuous(name="Petal Width (cm)", breaks = NULL, limits=c(1.0,2.5))

```

# 13. ggsave

------------------------------------------------------------------------

So far, these graphs have looked pretty underwhelming (not publication
quality).

On the R interface graphs do not look so great, but with ggsave, we can
save our graph as a jpg, pdf, etc...

Pick your favorite grpah from our exercises and see how it looks as a
saved file.

Look at how crisp that is!

```{r, echo=F, message=F, warning=F}
####
#13#
####

iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color=Species)) +
  geom_point() +
  labs(x = "Petal Length", y = "Petal Width", 
       title = "Petal Width vs. Petal Length") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_stata() +
  scale_color_brewer("Species", palette="Dark2") +
  scale_x_continuous (name= "Petal Length (cm)", breaks=seq(3.0, 7.0, 2.0),limits = c(3,7)) +
  scale_y_continuous(name="Petal Width (cm)", breaks = seq(1.0, 2.5, 0.25), limits=c(1.0,2.5))



ggsave("021124_First_Graph.jpg")
```

------------------------------------------------------------------------

# Appendix

```{r show-code, ref.label = all_labels(), echo = TRUE, eval = FALSE}

```
