---
title: "CM515 Day 2 Plotting with ggPlot Empty Workbook"
author: "Alex Alon"
date: "2024-02-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "CM515 Day 2: Plotting with ggplot()"
output: pdf_document
---

```{r setup, include=FALSE}
library(datasets)
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggthemes)




```

# 1. Putting it all together
*******

We are going to take some of my data and combine the principles from day1 of the ggplot section to make an awesome graph!

Context:

I have data for a bunch of strains where I measured the growth (as optical density) every hour for 26 hours in quadruplicate.

This experiment was done at 85C and 95C

The strains are:

TS559-Parent Strain 
TK0887-deleted for Endonuclease Q 
TK1898-deleted for Endonuclease MS 
TK2143-deleted for Uracil DNA Glycosylase
TK0887/TK1898 and TK1898/TK2143-deleted for both genes 
TK0887/TK1898/TK2143-deleted for all three genes




```{r, echo=T, message=F, warning=F}
####
#01#
####


dU_GC<-read.csv("Growth_Curve_dU.csv",fill=TRUE,header=TRUE)

head(dU_GC)


```

# 2. What happens when you use not tidy data.

*******

This data is not tidy, which means that we will have a lot lot lot of issues. Can we even graph this?

Try graphing it.


```{r, echo=T, message=F, warning=F}
####
#02#
####



```


# 2. What happens when you do statistics on not tidy data

I want to graph the average ODs for each strain at every hour. 

If I am using the current data in not tidy format, this is the amount of statistical analysis I have to do:

1. Pivot the data longer and make individual tables for each strain that summarize the data. I basically have to make tidy data using R which, while doable, is a lot of unnecessary work.

2. I then have to combine every single table for each individual strain together using rbind.



```{r, echo=T, message=F, warning=F}
####
#02#
####


GC_85_Blank <-dU_GC %>%
pivot_longer(cols = starts_with("Blank"), 
               names_to = "Blank Replicates", 
               values_to = "Blank Absorbance",
               values_drop_na = TRUE,
               values_transform = list(Absorbance=as.character)) %>%
  select(c(Hours, `Blank Replicates`, `Blank Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`Blank Absorbance`), sd= sd(`Blank Absorbance`)) %>%
  add_column(Strain = "Blank")

GC_85_TS559 <- dU_GC %>%
  pivot_longer(cols = starts_with("TS559"), 
               names_to = "TS559 Replicates", 
               values_to = "TS559 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `TS559 Replicates`, `TS559 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`TS559 Absorbance`), sd = sd(`TS559 Absorbance`)) %>%
  add_column(Strain = "TS559")

GC_85_0887 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK0887"), 
               names_to = "0887 Replicates", 
               values_to = "0887 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `0887 Replicates`, `0887 Absorbance`))%>%
  group_by(Hours) %>%
  summarize(mean = mean(`0887 Absorbance`), sd = sd(`0887 Absorbance`))%>%
  add_column(Strain = "TK0887-Endonuclease Q")


GC_85_1898 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK1898"), 
               names_to = "1898 Replicates", 
               values_to = "1898 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `1898 Replicates`, `1898 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`1898 Absorbance`), sd = sd(`1898 Absorbance`))%>%
  add_column(Strain = "TK1898-Endonuclease MS")

GC_85_2143 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK2143"), 
               names_to = "2143 Replicates", 
               values_to = "2143 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `2143 Replicates`, `2143 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`2143 Absorbance`), sd = sd(`2143 Absorbance`))%>%
  add_column(Strain = "TK2143-Uracil DNA Glycosylase")

GC_85_0887_1898 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK0887.TK1898"), 
               names_to = "TK0887/TK1898 Replicates", 
               values_to = "TK0887/TK1898 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `TK0887/TK1898 Replicates`, `TK0887/TK1898 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`TK0887/TK1898 Absorbance`), 
            sd = sd(`TK0887/TK1898 Absorbance`))%>%
  add_column(Strain = "TK0887/TK1898")

GC_85_0887_2143 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK0887.TK2143"), 
               names_to = "TK0887/TK2143 Replicates", 
               values_to = "TK0887/TK2143 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `TK0887/TK2143 Replicates`, `TK0887/TK2143 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`TK0887/TK2143 Absorbance`), 
            sd = sd(`TK0887/TK2143 Absorbance`))%>%
  add_column(Strain = "TK0887/TK2143")

GC_85_1898_2143 <- dU_GC %>%
  pivot_longer(cols = starts_with("TK1898.TK2143"), 
               names_to = "TK1898/TK2143 Replicates", 
               values_to = "TK1898/TK2143 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `TK1898/TK2143 Replicates`, `TK1898/TK2143 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`TK1898/TK2143 Absorbance`), 
            sd = sd(`TK1898/TK2143 Absorbance`))%>%
  add_column(Strain = "TK1898/TK2143")

GC_85_0887_1898_2143<- dU_GC %>%
  pivot_longer(cols = starts_with("TK0887.TK1898.TK2143"), 
               names_to = "TK0887/TK1898/TK2143 Replicates", 
               values_to = "TK0887/TK1898/TK2143 Absorbance",
               values_drop_na = TRUE) %>%
  select(c(Hours, `TK0887/TK1898/TK2143 Replicates`,
           `TK0887/TK1898/TK2143 Absorbance`)) %>%
  group_by(Hours) %>%
  summarize(mean = mean(`TK0887/TK1898/TK2143 Absorbance`), 
            sd = sd(`TK0887/TK1898/TK2143 Absorbance`))%>%
  add_column(Strain = "TK0887/TK1898/TK2143")


# combine tables
dU_Table <-rbind(GC_85_Blank, GC_85_TS559)
dU_Table <-rbind(dU_Table, GC_85_0887)
dU_Table <-rbind(dU_Table, GC_85_1898)
dU_Table <-rbind(dU_Table, GC_85_2143)
dU_Table <-rbind(dU_Table, GC_85_0887_1898)
dU_Table <-rbind(dU_Table, GC_85_0887_2143)
dU_Table <-rbind(dU_Table, GC_85_1898_2143)
dU_Table_Final <-rbind(dU_Table, GC_85_0887_1898_2143)


dU_Table_Final %>%
  ggplot(aes(x=Hours, y=mean, color=Strain)) +
  geom_line()
```



Not only is that an insane amount of work (took me several days to figure out) and a lot of lines of code, the graph is factually incorrect.

The lines of each strain are not exactly correct.

Because TK0887 is a repeating column, the 'startswith' argument cannot differentiate between each column that has the name TK0887.

This means way way more work, more time being frustrated, and less time making your graph awesome!

Let's make the data tidy, and then start over.

# 3. Working with tidy data. Mapping Aesthetics

We can complete the same statistical anlysis in a SIGNIFICANTLY shorter time span.

Try plotting mean (optical density) against time now, and factor the color by strain.

```{r, echo=T, message=F, warning=F}
####
#03#
####

dU_GC_tidy<-read.csv("Growth_Curve_dU_tidy.csv",fill=TRUE,header=TRUE)

head(dU_GC_tidy)

dU_GC_tidy_Stat <- dU_GC_tidy %>%
  group_by(Hours, Strain) %>%
  summarize(`Optical Density` = mean(OD), sd = sd(OD), n=n(), se = sd/sqrt(n)) %>%
  group_by(Strain)


```
# 4. Factoring the Mapping

Our legend labels do not say in any way that these are gene deletions. I did not include deltas because R would not interpret this from a csv file.

I recommend always fixing the names of columns, Rows, and categorical variables before graphing because afterwards it is a lot of code to fix those issues.

We should put a delta sign in front of each gene number, but how?

Fortunately, R can interpret unicode.

Using labels, we can rename our legend to include deltas in front of each gene.

https://en.wikipedia.org/wiki/List_of_Unicode_characters 

You can find all greek letters on wikipedia using the link above under Greek and Coptic, or just ask ChatGPT. We put the unicode right into the string (unicode for delta is \u0394).




```{r, echo=T, message=F, warning=F}
####
#04#
####


dU_GC_tidy_Stat <- dU_GC_tidy_Stat %>%
  mutate(Strain = 
           recode(Strain, "TK0887-Endonuclease Q" = "\u0394TK0887-Endonuclease Q", 
                  "TK1898-Endonuclease MS" = "\u0394TK1898-Endonuclease MS", 
                  "TK2143-Uracil DNA Glycosylase" = "\u0394TK2143-Uracil DNA Glycosylase",
                  "TK0887/TK1898" = "\u0394TK0887/\u0394TK1898", 
                  "TK0887/TK2143" = "\u0394TK0887/\u0394TK2143", 
                  "TK1898/TK2143" = "\u0394TK1898/\u0394TK2143", 
                  "TK0887/TK1898/TK2143" = "\u0394TK0887/\u0394TK1898/\u0394TK2143"))

#now let's see how the graph looks with the mutated data

  

```


# 5. Picking the right geometry

Now the data has received proper statistics and been mapped.

We can now tell the dots apart, but, maybe a scatter plot is not the best way to represent this data.

Try a few geometries and note the pros/cons.



```{r, echo=T, message=F, warning=F}
####
#05#
####


```

# 5b. Overlapping Geometries

Try combining a line graph and scatter plot. Adjust the size of the points and the lines so that you can both distinctly



```{r, echo=T, message=F, warning=F}
####
#05b#
####




```

# 6. Labels

Now this graph is starting to display our data appropriately.

Let's add axis labels.

We want to specify that the y axis is optical density at 600nm.

X axis is Time in hours

Bquote allows us to customize our text to be bold, italicized and underlined. We can also add superscripts, subscripts, and many other components.

For more on bquote-use help("plotmath") or ask ChatGPT.

```{r, echo=T, message=F, warning=F}
####
#06#
####

#Try to add your own labels and compare how they look with the bquote adjustments below.

dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  labs(title = "Strains lacking all dU repair components have the most severe fitness defect", subtitle="85\u00B0C",
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (hrs)'))) +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle= element_text(hjust = 0.5)) 
  

```

# 7. Scales

The scales are currently not very informative.
Lets rescale the x axis tick marks to be more helpful. Try to get it to have a tick mark every 4 hours.
Try also rescaling the y axis so we have context for where the maximum OD is by 0.3.


```{r, echo=T, message=F, warning=F}
####
#07#
####

                                                                

```


# 8. Scales-Colors

Right now the legend is pretty out of order and the rainbow of colors is hard to follow.

Let's make a color palette we can work with using scale_colors. You decide if you want manual or brewer.


```{r, echo=T, message=F, warning=F}
####
#08#
####




```
# 8b. Set your colors to your legend

Now we have our own colors, but the legend is not in any logical order!

As with scale_x_continuous, we can also add breaks to scale_color_manual to reorder our legend names to our colors.


```{r, echo=T, message=F, warning=F}
####
#08b#
####

dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  labs(title = bquote(bold("Strains lacking all dU repair components have the most severe fitness defect")), subtitle="85\u00B0C",
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle= element_text(hjust = 0.5)) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.2), expand = c(0,0)) +
  scale_color_manual(values=c("pink", "brown4", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black" ), name = "Strain", breaks=c("Blank", "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143")) +
  geom_errorbar(width=0.1, aes(x=Hours, ymax = mean+sd, ymin = mean-sd, color=Strain)) 

 
```


# 9. Theme

Now that we are done graphing, choose your favorite theme.

Sometimes, if you put a theme, you may want to put it early in your ggplot code. Otherwise the theme is building off of every previous component and this can cause errors.


```{r, echo=T, message=F, warning=F}
####
#9#
####


  



```

# 10. ggSave

It looks like we are ready to save the graph!

Make any changes you want that you think make the scales and colors look best.

Then save your graph.

```{r, echo=T, message=F, warning=F}
####
#10#
####


ggsave("Growth Curve.jpg")
 
```
# 11. Facet Wrapping

BUT WAIT! THERE IS MORE!

We never did figure out the issue of overlapping graphs after all.

We can separate the graphs into smaller individual graphs for each strain.

```{r, echo=T, message=F, warning=F}
####
#11#
####
dU_GC_tidy<-read.csv("Growth_Curve_dU_tidy.csv",fill=TRUE,header=TRUE)

dU_Neg_85 <- dU_GC_tidy %>%
  filter(Strain == "Blank") %>%
  group_by(Hours)%>%
  summarize(blank_mean = mean(OD), .groups = "drop")

dU_GC_tidy_Stat<- dU_GC_tidy %>%
  left_join(dU_Neg_85, by = "Hours") %>%
  mutate(OD = OD - blank_mean) %>%
  group_by(Hours, Strain) %>%
  summarize(mean = mean(OD), sd = sd(OD), n = n(), se = sd / sqrt(n), .groups = "drop") %>%
  filter(Strain != "Blank") %>%
  mutate(Strain = 
           recode(Strain, "TK0887-Endonuclease Q" = "\u0394TK0887-Endonuclease Q", 
                  "TK1898-Endonuclease MS" = "\u0394TK1898-Endonuclease MS", 
                  "TK2143-Uracil DNA Glycosylase" = "\u0394TK2143-Uracil DNA Glycosylase",
                  "TK0887/TK1898" = "\u0394TK0887/\u0394TK1898", 
                  "TK0887/TK2143" = "\u0394TK0887/\u0394TK2143", 
                  "TK1898/TK2143" = "\u0394TK1898/\u0394TK2143", 
                  "TK0887/TK1898/TK2143" = "\u0394TK0887/\u0394TK1898/\u0394TK2143"))
                                                 
dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  theme_bw() +
  labs(title = bquote(bold("Strains lacking all dU repair components have the most severe fitness defect")), subtitle=bquote(bold("85\u00B0C")),
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.2), expand = c(0,0)) +
  scale_color_manual(values=c( "darkgrey", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black"), name = "Strain", breaks=c("TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))+
  facet_wrap(~Strain) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 12), plot.subtitle= element_text(hjust =   0.5, size = 10)) +
    theme(axis.text.x = element_text(face = 'bold', size = 8, color = "black")) +
    theme(axis.text.y = element_text(face = 'bold', size = 8, color = 'black')) +
    theme(axis.title.x = element_text(face = 'bold', size = 12)) +
    theme(axis.title.y = element_text(face = 'bold', size = 12)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.text = element_text(face = 'bold', size = 8)) +
    theme(legend.title = element_text( face = 'bold', size = 10)) 

ggsave("dU_Plot_85_Facet_Wrapped.jpg", dpi=2000)

#facet_wrap(~factor(Strain, levels=c( "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))) +

```

# 11b. Facet-Wrap - Removing Legends

It is a bit redundant to have the panel labels and the strain labels.

We can remove one or the other.

```{r, echo=T, message=F, warning=F}
####
#11b#
####


#without the legend


dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  theme_bw() +
  labs(title = bquote(bold("Strains lacking all dU repair components have the most severe fitness defect")), subtitle=bquote(bold("85\u00B0C")),
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.2), expand = c(0,0)) +
  scale_color_manual(values=c( "darkgrey", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black"), name = "Strain", breaks=c("TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))+
  facet_wrap(~factor(Strain, levels=c( "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 12), plot.subtitle= element_text(hjust =   0.5, size = 10)) +
    theme(axis.text.x = element_text(face = 'bold', size = 8, color = "black")) +
    theme(axis.text.y = element_text(face = 'bold', size = 8, color = 'black')) +
    theme(axis.title.x = element_text(face = 'bold', size = 12)) +
    theme(axis.title.y = element_text(face = 'bold', size = 12)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.text = element_text(face = 'bold', size = 8)) +
    theme(legend.title = element_text( face = 'bold', size = 10)) +
    theme(strip.text = element_text(face = "bold", size=6))+
    theme(legend.position = "none") 

ggsave("dU Growth Curve_Facet_Wrapped.jpg", dpi = 2000)


#without the panel labels


dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  theme_bw() +
  labs(title = bquote(bold("Strains lacking all dU repair components have the most severe fitness defect")), subtitle=bquote(bold("85\u00B0C")),
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.2), expand = c(0,0)) +
  scale_color_manual(values=c( "darkgrey", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black"), name = "Strain", breaks=c("TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))+
  facet_wrap(~factor(Strain, levels=c( "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 12), plot.subtitle= element_text(hjust =   0.5, size = 10)) +
    theme(axis.text.x = element_text(face = 'bold', size = 8, color = "black")) +
    theme(axis.text.y = element_text(face = 'bold', size = 8, color = 'black')) +
    theme(axis.title.x = element_text(face = 'bold', size = 12)) +
    theme(axis.title.y = element_text(face = 'bold', size = 12)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.text = element_text(face = 'bold', size = 8)) +
    theme(legend.title = element_text( face = 'bold', size = 10)) +
    theme( strip.text.x = element_blank() )

ggsave("dU Growth Curve_Facet_Wrapped_2.jpg", dpi=2000, width = 12)
 
```

# 12. Bonus- ggpubr

There are a TON of packages that can take ggplot to the next level.

I actually collected data for both 85 degrees and 95 degrees C.

If I want, I can use the ggpubr package to get an image with two graphs.

```{r, echo=T, message=F, warning=F}
####
#12#
####

install.packages("ggpubr")
library(ggpubr)
#make 85 degree data into an item

dU_85 <- dU_GC_tidy_Stat%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  theme_bw() +
  labs(subtitle=bquote(bold("85\u00B0C")),
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.4), expand = c(0,0)) +
  scale_color_manual(values=c( "darkgrey", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black"), name = "Strain", breaks=c("TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))+
  facet_wrap(ncol=8, ~factor(Strain, levels=c( "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 12), plot.subtitle= element_text(hjust =   0.5, size = 10)) +
    theme(axis.text.x = element_text(face = 'bold', size = 8, color = "black")) +
    theme(axis.text.y = element_text(face = 'bold', size = 8, color = 'black')) +
    theme(axis.title.x = element_text(face = 'bold', size = 12)) +
    theme(axis.title.y = element_text(face = 'bold', size = 12)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.text = element_text(face = 'bold', size = 8)) +
    theme(legend.title = element_text( face = 'bold', size = 10)) +
    theme(strip.text = element_text(face = "bold", size=4))+
    theme(legend.position = "none") 

dU_85


#95 degree data

dU_GC_tidy_95<-read.csv("Growth_Curve_dU_tidy_95.csv",fill=TRUE,header=TRUE)

dU_Neg_95 <- dU_GC_tidy_95 %>%
  filter(Strain == "Blank") %>%
  group_by(Hours)%>%
  summarize(blank_mean = mean(OD), .groups = "drop")

dU_GC_tidy_Stat_95<- dU_GC_tidy_95 %>%
  left_join(dU_Neg_95, by = "Hours") %>%
  mutate(OD = OD - blank_mean) %>%
  group_by(Hours, Strain) %>%
  summarize(mean = mean(OD), sd = sd(OD), n = n(), se = sd / sqrt(n), .groups = "drop") %>%
  filter(Strain != "Blank") %>%
  mutate(Strain = 
           recode(Strain, "TK0887-Endonuclease Q" = "\u0394TK0887-Endonuclease Q", 
                  "TK1898-Endonuclease MS" = "\u0394TK1898-Endonuclease MS", 
                  "TK2143-Uracil DNA Glycosylase" = "\u0394TK2143-Uracil DNA Glycosylase",
                  "TK0887/TK1898" = "\u0394TK0887/\u0394TK1898", 
                  "TK0887/TK2143" = "\u0394TK0887/\u0394TK2143", 
                  "TK1898/TK2143" = "\u0394TK1898/\u0394TK2143", 
                  "TK0887/TK1898/TK2143" = "\u0394TK0887/\u0394TK1898/\u0394TK2143"))

dU_95 <- dU_GC_tidy_Stat_95%>%
  ggplot(aes(x=Hours, y=mean, color=Strain, ymax=mean+se, ymin=mean-se)) +
  geom_line(size=0.5) +
  geom_point(size=1.2) +
  geom_errorbar(width=0.1) +
  theme_bw() +
  labs( subtitle=bquote(bold("95\u00B0C")),
       y = bquote(bold("OD")[600][nm]),
       x = bquote(bold('Time (Hours)'))) +
  scale_x_continuous(breaks=c(0, 4, 8, 12, 16, 20, 24), limits=c(0,26), expand = c(0,0)) +
  scale_y_continuous(breaks=c(0, 0.3, 0.6, 0.9, 1.2), limits=c(-0.05, 1.4), expand = c(0,0)) +
  scale_color_manual(values=c( "darkgrey", "red3", "yellow3", "steelblue3", "orange2", "purple3", "green2", "black"), name = "Strain", breaks=c("TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))+
  facet_wrap(ncol=8, ~factor(Strain, levels=c( "TS559-Parent Strain", "\u0394TK0887-Endonuclease Q", "\u0394TK1898-Endonuclease MS", "\u0394TK2143-Uracil DNA Glycosylase","\u0394TK0887/\u0394TK1898", "\u0394TK0887/\u0394TK2143", "\u0394TK1898/\u0394TK2143", "\u0394TK0887/\u0394TK1898/\u0394TK2143"))) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 12), plot.subtitle= element_text(hjust =   0.5, size = 10)) +
    theme(axis.text.x = element_text(face = 'bold', size = 8, color = "black")) +
    theme(axis.text.y = element_text(face = 'bold', size = 8, color = 'black')) +
    theme(axis.title.x = element_text(face = 'bold', size = 12)) +
    theme(axis.title.y = element_text(face = 'bold', size = 12)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.text = element_text(face = 'bold', size = 8)) +
    theme(legend.title = element_text( face = 'bold', size = 10)) +
    theme(strip.text = element_text(face = "bold", size=4))+
    theme(legend.position = "none") 
dU_95
#ggarange

dU_Plot <- ggarrange(dU_85, dU_95, nrow = 2, legend = "right")

dU_Plot <-annotate_figure(dU_Plot, top = text_grob("Strains lacking all dU repair components have the most severe fitness defect", 
               color = "black", face = "bold", size = 14))

dU_Plot

ggsave("dU_GC_Final.jpg", dU_Plot, dpi=2000, width = 10, device = 'jpeg')


```

# Appendix

```{r show-code, ref.label = all_labels(), echo = TRUE, eval = FALSE}

```
